Acceleration of plague outbreaks in the second pandemic

### 2020-07-05 ms_text.tex
minor changes during 'final' proofread

We can use independent estimates of the generation interval to estimate $\rzero$ (and hence the theoretical attack rate) for London epidemics under different transmission modes. By considering the additional (extreme) assumptions that \hypertarget{4assumptions}{} (1) the population was homogeneously mixed; (2) the population was initially 100\% susceptible; (3) there were no asymptomatic infections; and (4) all infected individuals died (\ie the CFP was 100\%), all of which increase the observed attack rate, we can calculate an upper bound on the total mortality we would expect to observe in an epidemic with a particular transmission mode.

@@

As discussed above, we initially fitted a Richards model, which extends the logistic equation with a shape parameter, but found that we could not get stable fits to the smallest/noisiest epidemics; we never found qualitative differences between growth rates estimated from logistic and Richards fits.  Ma \etal found that a \emph{delayed} logistic curve, which adds an exponentially distributed delay between infection and mortality, estimated growth rates from mortality curves with less bias. However, the delay between infection and mortality in our data is extremely uncertain---it would represent another parameter that would have to be estimated from noisy data. For wills data, the delay could even be negative, \ie wills could have been written out of fear of \emph{future} infection. Furthermore, we found little difference between the Richards and logistic fits---which Ma \etal found to have opposite biases; hence these methods are likely to bracket the growth rate estimated by the delayed logistic.

@@

The fitting windows used for all our fits are listed in \supp \cref{tab:windows,tab:windowsminor}. Although we experimented with different methods for determining the fitting window (e.g. using a threshold of 1/100 of the peak mortality rate, or using the observation after the last local minimum before the peak), 
we emphasize that our choices were entirely determined by seeking ``good'' fits (judged by visual inspection), and not by the pattern of estimated initial growth rates. Moreover, our conclusion that later plagues were much faster than the earlier plagues is robust to choices we made about window selection.

>>>>

We can use independent estimates of the generation interval to estimate $\rzero$ (and hence the theoretical attack rate) for London epidemics under different transmission modes. By considering the additional (extreme) assumptions that \hypertarget{4assumptions}{} (1) the population was homogeneously mixed; (2) the population was initially 100\% susceptible; (3) there were no asymptomatic infections; and (4) all infected individuals died (\ie the CFP was 100\%), all of which increase the observed attack rate, we can calculate an upper bound on the total mortality we would expect to observe in an epidemic with a particular transmission mode.
%%
\david{I meant that if CFP$=1$ then all infected individuals get
  detected by mortality statistics, hence this would yield larger
  observed attack rate than if CFP$<1$.  Not sure this is
  sufficiently clear.}

@@

As discussed above, we initially fitted a Richards model, which extends the logistic equation with a shape parameter, but found that we could not get stable fits to the smallest/noisiest epidemics; we never found qualitative differences between growth rates estimated from logistic and Richards fits.  Ma \etal found that a \emph{delayed} logistic curve, which adds an exponentially distributed delay between infection and mortality, estimated growth rates from mortality curves with less bias. However, the delay between infection and mortality in our data is extremely uncertain---it would represent another parameter that would have to be estimated from noisy data. For wills data, the delay could even be negative, \ie wills could have been written out of fear of \emph{future} infection. Furthermore, because we found little difference between the Richards and logistic fits---which Ma \etal found to have opposite biases---growth rates estimated by these two methods are likely to bracket the growth rate estimated by the delayed logistic.

@@

The fitting windows used for all our fits are listed in \supp \cref{tab:windows,tab:windowsminor}. Although we experimented with different methods for determining the fitting window (e.g., starting the window from the observation after the last local minimum before the peak, or based on a threshold of $1/100$ of the peak mortality rate), we emphasize that our choices were entirely determined by seeking ``good'' fits (judged by visual inspection), and not by the pattern of estimated initial growth rates. Moreover, our conclusion that later plagues were much faster than the earlier plagues is robust to choices we made about window selection.
